Inferring stellar rotation periods using Gaussian processes

### 2016-12-24 documents/ms.tex

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}

@@

In order to manage the extreme flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any particular
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of 
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors.
% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential 
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior 
 and use a mix of different proposal distributions: 
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves. 
[DFM: more words/citations here?]  
We run fifty steps of the sampler at a time, checking for convergence after 
each iteration, up to a maximum of fifty iterations.  We declare convergence if 
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(7.2, 1.2)$ & (2, 20) \\
    $\ln \Gamma$ & $\mathcal N(-2.3, 1.4)$ & (-10, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform & ($\ln 0.5, \ln 100$) \\ 
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}

@@

Figure \ref{fig:compare_mcmc} summarizes the results of this MCMC fitting procedure
compared to the injected ``true'' stellar rotation periods. 
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

>>>>

\documentclass[useAMS, usenatbib, preprint, 12pt]{aastex}
\usepackage{cite, natbib}
\usepackage{float}
\usepackage{epsfig}
\usepackage{cases}
\usepackage[section]{placeins}
\usepackage{graphicx, subfigure}
\usepackage{color}
\usepackage{bm}
\usepackage{enumerate}

@@

In order to manage the great flexibilty of the GP model and focus on 
reliably retrieving the period parameter, we also impose priors
on the non-period GP parameters that we 
learned during the development of this method.  
Beginning with very broad priors (uniform in log space between 
-20 and 20) on all the parameters, we first explored what the 
typical values of the non-period parameters tended to be when the 
injected periods were successfully recovered.  We also experimented with
constraining the allowed ranges of the parameters, after discovering 
that some regions of parameter space (such as large values of $A$ and $\Gamma$
and small values of $l$) tended to allow fits that ignored the desired
periodicity.  The priors and bounds we used for this work are listed in
table \ref{tab:priors}.  We note that we do not have any specific
quantitative justification for the use of these priors,
except that they produce good results for this particular test set of 
light curves.  We thus caution that other data sets might have different
noise properties, potentially requiring modifications to these priors
on the non-period parameters.  For period, we tried two different methods:
an uninformed (log-flat) prior between 0.5\,d and 100\,d, and an ACF-informed
prior (details in Appendix.)

% We note that our default prior for period is 
% log-uniform between the bounds, but that we were also able to 
% further improve our results by using an ACF-inspired prior.

To sample the posterior in a way that is sensitive to potential 
multimodality, we use the \texttt{emcee3} MCMC sampler.
We initialize 500 walkers with random samples from the prior 
 and use a mix of different proposal distributions: 
specifically, a 2:2:1 mix of KDE, differential evolution, and DESnooker moves. 
[DFM: more words/citations here?]  
We run fifty steps of the sampler at a time, checking for convergence after 
each iteration, up to a maximum of fifty iterations.  We declare convergence if 
the total effective chain length is at least 8$\times$ the maximum autocorrelation
time.  When convergence is achieved, we drop the first two autocorrelation 
lengths in the chain as a burn-in, and randomly choose 5000 samples as 
representative of the posterior.  This fitting process takes several hours 
for a typical simulated light curve, though in some cases it may take 12
hours or longer to converge.

\begin{table*}
\begin{center}
\caption{Priors and bounds on the natural logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Prior & Bounds\\
    \hline
    $\ln A$ & $\mathcal N(-13, 5.7)$ & (-20, 0) \\
    $\ln l$ & $\mathcal N(7.2, 1.2)$ & (2, 20) \\
    $\ln \Gamma$ & $\mathcal N(-2.3, 1.4)$ & (-10, 3) \\
    $\ln \sigma$ & $\mathcal N(-17, 5)$ & (-20, 0) \\
    $\ln P $ & Uniform / ACF-based & ($\ln 0.5, \ln 100$) \\ 
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}

@@

Figures \ref{fig:compare_mcmc_noprior} and \ref{fig:compare_mcmc_acfprior} 
summarize the results of this MCMC fitting 
procedure compared to the injected ``true'' stellar rotation periods, for the 
uninformed and ACF-informed priors on period, respectively.
The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

@@

\appendix

\section{An ACF-informed Prior for Rotation Period}

The autocorrelation function (ACF) has proven to be very useful
for measuring the rotation periods of stars from time-series
photometry.  
However, as discussed in the body of this paper, the
ACF-based method has several shortcomings, 
most notably the inability to deliver uncertainties, but also 
the necessity of several heuristic choices, 
such as a timescale on which to smooth the ACF, 
how to define a peak, whether the first or second peak 
gets selected, and what constitutes a secure detection.
While this paper presents a new MCMC-based method
that avoids these shortcomings, 
it seems prudent to still use information available from the ACF.
This Appendix describes how we use the ACF to define a prior on period.

Our strategy here is not to attempt to decide which single
peak in the ACF best represents the true rotation period,
but rather to identify several \emph{candidate} periods and define
a weighting scheme in order to create a non-commital, though useful, 
multimodal prior.  While this procedure does not avoid heuristic choices, 
the fact that its goal is simply to create a \emph{prior} for a 
probabilitistically justified inference scheme rather than
to identify a single correct period softens the potential impact 
of these choices.

As another innovation beyond what ACF methods in the literature have 
presented, we also bandpass filter the light curves (using a 5th order
Butterworth filter, as implemented in \texttt{SciPy}) before calculating the 
autocorrelation.  This suppresses power on timescales shorter than a chosen 
minimum period $P_{\rm min}$ and longer than a chosen maximum $P_{\rm max}$,
producing a cleaner autocorrelation signal than an unfiltered light curve.

We use the following procedure to construct a prior for rotation period
given a light curve:
\begin{enumerate}
\item{For each value of $P_i$, where $i = \{1, 3, 5, 10, 30, 50, 100\}$\,d,
we apply a bandpass filter to the light curve using $P_{\rm min}=0.1$\,d
and $P_{\rm max} = P_i$.  We then calculate the ACF of the filtered 
light curve out to a maximum lag of $2P_i$ and smooth it with a boxcar 
filter of width $P_i/10$.}

\item{We identify the time lag corresponding to the 
first peak of each of these ACFs, as well as the first peak's
trough-to-peak height, creating a set of candidate periods
$T_i$ and heights $h_i$.}

\item{We assign a quality metric $Q_i$ to each of these candidate
periods, as follows.  First, we model the ACF as a 
damped oscillator with fixed period $T_i$:
\begin{equation}
y = A e^{-t/\tau} \cos{\frac{2\pi t}{T_i} },
\end{equation}
where and $t$ is the lag time, 
and find the best-fitting parameters $A_i$ and $\tau_i$ by a non-linear
least squares minimization procedure.  We then define the
following heuristic quality metric:
\begin{equation}
\label{eq:quality}
Q_i = \left(\frac{\tau_i}{T_i}\right) \left(\frac{N_i h_i}{R_i}\right),
\end{equation}
where $h_i$ is the height of the ACF peak at $T_i$, 
$N_i$ is the length of the lag vector in the ACF (directly proportional
to the maximum allowed period $P_i$), 
and $R_i$ is the sum of squared residuals between the 
damped oscillator model and the actual ACF data.  The idea behind this 
quality metric to give a candidate period a higher score if

    \begin{enumerate}[(a)]

    \item{it has many regular sinusoidal peaks, such that the decay
        time $\tau_i$ is long compared to the oscillation period $T_i$,}
    \item{the ACF peak height is high, and}
    \item{the damped oscillator model is a good fit (in a $\chi^2$
      sense) to the ACF, with extra bonus for being 
      a good fit over more points (larger $N_i$, or longer $P_i$).}

    \end{enumerate}
}

\item{Given this set of candidate periods $T_i$ and quality metrics $Q_i$,
we finally construct a multimodal prior on the $P$ parameter of the GP
model as a weighted mixture of Gaussians:
\begin{equation}
\label{eq:mixture}
p(\ln P) = \frac {\displaystyle \sum_i Q_i \left(0.9\mathcal N(\ln T_i, 0.2) + 
                                          0.05\mathcal N(\ln (T_i/2), 0.2) +
                                          0.05\mathcal N(\ln (2 T_i), 0.2) \right)}
                {\sum_i Q_i}.
\end{equation}
That is, in addition to taking the candidate periods themselves as mixture
components, we also mix in twice and half each candidate period at a lower level, 
which compensates for the possibility that the first peak in the ACF may actually 
represent half or twice the actual rotation period.  The period width of 0.2 in 
log space (corresponding to roughly 20\% uncertainty) is again a heuristic choice,
balancing a healthy specificity with the desire to not have the results of 
the inference overly determined by the ACF prior.
}

Incidentally, while we use the procedure described here to create a prior on 
$P$ which we use while inferring the parameters of the quasi-periodic GP model,
this same procedure may also be used in the service of a
rotation-period estimating procedure all on its own, perhaps being even more 
robust and accurate than the traditional ACF method.  We leave exploration of 
this possibility to future work.

\end{enumerate}
